---
title: "HRP 203 Final Project"
author: "Genna Campain"
date: "5/28/24"
format:
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
---

## Assignment Description

Create a 4 to 6 page Quarto reproducible document to synthesize concepts learned throughout the course and particularly from Module 3. Create the report in a repository on your GitHub account based on the cohort simulated data. Include the following features: introduction section, methods section with notation, results section with data summary table and at least 2 figures, discussion. The assignment is due at 11am Pacific time on Monday, June 10th.

## Introduction

## Methods

For this project, I will be using a variety of packages which I load in the code chunk below.

```{r}
#| message: false
#| warning: false
library(dplyr)
# skewness function
library(e1071)
# coefficient plots
library(coefplot)
```

I will be using the `cohort` simulated data set provided through the course GitHub repository. A copy of this data set can be found in the `Data` folder in the GitHub repository for this project.

## Results

First,

### Data Summary Table

### Figure 1

### Figure 2

## Discussion

```{r}
1 + 1
```
